---
title: "TP2 - MATH60633A.H2026 - Statistical Methods for Financial Data "
author: "Francis Mazzone () & Vaast Émilien (11364986)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document presents the analysis of the data collected for our project. We will explore the data, perform statistical tests, and visualize the results to draw meaningful conclusions.

```{r libraries, message=FALSE, echo= TRUE}
# Load necessary libraries
library(quantmod)
library(knitr)
#install.packages("patchwork")
library(patchwork)
#install.packages("kableExtra")
library(kableExtra)
#install.packages("here")
library(here)
library(xts)
#install.packages("PerformanceAnalytics")
library(PerformanceAnalytics)
library(zoo)
#install.packages("dplyr")
#install.packages("lubridate")
library(dplyr)
library(lubridate)
library(ggplot2)
library(devtools)
library(tidyverse)
```

```{r environment set up, message=FALSE, warning=FALSE}
data_env <- new.env()
source(here("Functions", "f_black_scholes.R"))
source(here("Functions", "f_normal_scenarios.R"))

file_path <- here("Data", "Raw", "Market.rda")
stopifnot(file.exists(file_path))

load(file_path, envir = data_env)
ls(data_env)
(data <- data_env$Market)

sp500 <- data$sp500
vix <- data$vix 

rf_structure <- data.frame(
  y_maturity = as.numeric(attr(data[["rf"]], "names")),
  rate = as.numeric(data[["rf"]])
)

calls <- data$calls
puts <- data$puts

```

## Pricing of a portfolio of options

```{r define conventions, message=FALSE, warning=FALSE}
# Define conventions
days_in_year <- 360
trading_days_in_year <- 250
```

We want to build a portfolio that consists of 4 European call options:

-   Strike price: 1600 \| Maturity: 20 days
-   Strike price: 1650 \| Maturity: 20 days
-   Strike price: 1750 \| Maturity: 40 days
-   Strike price: 1800 \| Maturity: 40 days

```{r Build the portfolio, message=FALSE, warning=FALSE}
#Underlying asset price and volatility for the options in the portfolio
last_trading_price <- coredata(tail(sp500, 1))
last_volatility <- coredata(tail(vix, 1))

# Define the portfolio with the specified options
maturities_day = c(20,20,40,40) 
strike_prices = c(1600, 1650, 1750, 1800)

#Build the portfolio data frame
portfolio <- data.frame(
  Maturity_days = maturities_day, 
  Strike_Price = strike_prices,
  Trading_Price = rep(last_trading_price, 4),
  Volatility = rep(last_volatility, 4)
)
```

```{r Calculate option prices, message=FALSE, warning=FALSE}
# Calculate the option prices for each option in the portfolio using the Black-Scholes formula
portfolio <- portfolio %>% 
  rowwise() %>%
  mutate(
    Call_Price = f_black_scholes(
      S    = Trading_Price,
      K    = Strike_Price,
      T    = Maturity_days / trading_days_in_year,
      trading_day_convention = trading_days_in_year,
      rate_day_convention = days_in_year,
      rf_structure = rf_structure,
      sigma = Volatility
    ) %>% filter(Option_Type == "Call") %>% pull(Price)
  ) %>%
  ungroup() 

portfolio
```

## One risk driver and uni-variate Gaussian Model
```{r}
#Computing the log-returns
daily_log_returns <- diff(log(sp500))[-1]
 
#Calculate the mean and standard deviation of the daily log-returns
#They serve as estimators for the parameters of the uni-variate Gaussian model
mu_daily_log_returns <- mean(daily_log_returns)
sigma_daily_log_returns <- sd(daily_log_returns)

# Generate 10,000 scenarios for the underlying asset price in 5 days using the uni-variate Gaussian model
days_ahead <- 5

scenarios <- f_normal_scenarios(
  mu_returns = mu_daily_log_returns,
  sigma_returns = sigma_daily_log_returns,
  n_scenarios = 10000,
  T = days_ahead,
  spot_price = last_trading_price
)
```


```{r}
#Determine the P&L distribution of the book of options, using the simulated underlying values. Assume
#the implied volatility stays the same. Take interpolated rates for the term structure.

portfolio_value_inception <- sum(portfolio$Call_Price)
# Calculate the option prices at the end of the 5 days for each scenario
portfolio_value_five_days_ahead <- sapply(scenarios, function(scenario_price) {
  sum(
    mapply(function(strike, maturity) {
      f_black_scholes(
        S = scenario_price,
        K = strike,
        T = (maturity - 5) / trading_days_in_year,
        rf_structure = rf_structure,
        sigma = last_volatility
      ) %>% filter(Option_Type == "Call") %>% pull(Price)
    }, portfolio$Strike_Price, portfolio$Maturity_days)
  )
})

# Calculate the P&L distribution
pnl_distribution <- portfolio_value_five_days_ahead - portfolio_value_inception

# Compute the VaR95 and ES95
VaR95 <- quantile(pnl_distribution, 0.05)
ES95  <- mean(pnl_distribution[pnl_distribution <= VaR95])

# Plot the P&L distribution
pnl_df <- data.frame(PnL = pnl_distribution)

ggplot(pnl_df, aes(x = PnL)) +
  geom_histogram(binwidth = 1, fill = "#2C3E50", color = "#AEB6BF", alpha = 0.85) +
  geom_vline(xintercept = VaR95, color = "cyan", linewidth = 0.9) +
  geom_vline(xintercept = ES95,  color = "darkblue", linewidth = 0.9) +
  annotate("text", x = VaR95, y = Inf, 
           label = paste0("VaR 95%: $", round(VaR95, 2)),
           color = "cyan", hjust = -0.1, vjust = 2, size = 3.5, fontface = "bold") +
  annotate("text", x = ES95, y = Inf,  
           label = paste0("ES 95%: $",  round(ES95, 2)),
           color = "darkblue", hjust = -0.1, vjust = 4, size = 3.5, fontface = "bold") +
  labs(title    = "P&L Distribution — Portfolio of European Call Options",
       subtitle = paste0("10,000 scenarios | 5-day horizon | ",
                         "VaR 95%: $", round(VaR95, 2), 
                         " | ES 95%: $", round(ES95, 2)),
       x = "P&L ($)", 
       y = "Frequency",
       caption = "Univariate Gaussian model — one risk driver") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "#566573", size = 10),
    plot.caption  = element_text(color = "#566573", size = 9, hjust = 0),
    panel.grid.minor = element_blank()
  )
  
  
```











