# ---
title: "TP2 - MATH60633A.H2026 - Statistical Methods for Financial Data "
author: "Francis Mazzone () & Vaast Émilien (11364986)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document presents the analysis of the data collected for our project. We will explore the data, perform statistical tests, and visualize the results to draw meaningful conclusions.

```{r libraries, message=FALSE, echo= TRUE}
# Load necessary libraries
library(knitr)
library(patchwork)
library(here)
library(xts)
library(dplyr)
library(purrr)
library(ggplot2)
library(MASS)
library(plotly)
```

```{r environment set up, message=FALSE, warning=FALSE}
data_env <- new.env()
source(here("Functions", "f_black_scholes.R"))
source(here("Functions", "f_normal_scenarios.R"))
source(here("Functions", "f_normal_scenarios_bivar.R"))
source(here("Functions", "f_pnl_univariate.R"))
source(here("Functions", "f_pnl_bivariate.R"))
source(here("Functions", "f_vol_surface_workflow.R"))

file_path <- here("Data", "Raw", "Market.rda")
stopifnot(file.exists(file_path))

load(file_path, envir = data_env)
ls(data_env)

data <- data_env$Market

sp500 <- data$sp500
vix <- data$vix 

rf_structure <- data.frame(
  y_maturity = as.numeric(attr(data[["rf"]], "names")),
  rate = as.numeric(data[["rf"]])
)

calls <- data$calls
puts <- data$puts

# check class of all 
data_classes <- sapply(data, class)
print(data_classes)

```

## Pricing of a portfolio of options

```{r define conventions, message=FALSE, warning=FALSE}
# Define conventions
days_in_year <- 360
trading_days_in_year <- 250
```

We want to build a portfolio that consists of 4 European call options:

-   Strike price: 1600 \| Maturity: 20 days
-   Strike price: 1650 \| Maturity: 20 days
-   Strike price: 1750 \| Maturity: 40 days
-   Strike price: 1800 \| Maturity: 40 days

```{r Build the portfolio, message=FALSE, warning=FALSE}
#Underlying asset price and volatility for the options in the portfolio
last_trading_price <- coredata(tail(sp500, 1))
last_volatility <- coredata(tail(vix, 1))

# Define the portfolio with the specified options
maturities_day = c(20,20,40,40) 
strike_prices = c(1600, 1650, 1750, 1800)

#Build the portfolio data frame
portfolio <- data.frame(
  Maturity_days = maturities_day, 
  Strike_Price = strike_prices,
  Trading_Price = rep(last_trading_price, 4),
  Volatility_last_vix = rep(last_volatility, 4)
)
```

```{r Calculate option prices, message=FALSE, warning=FALSE}
# Calculate the option prices for each option in the portfolio using the Black-Scholes formula
portfolio <- portfolio %>% 
  rowwise() %>%
  mutate(
    Call_Price = f_black_scholes(
      S    = Trading_Price,
      K    = Strike_Price,
      T    = Maturity_days / trading_days_in_year,
      trading_day_convention = trading_days_in_year,
      rate_day_convention = days_in_year,
      rf_structure = rf_structure,
      sigma = Volatility_last_vix
    ) %>% filter(Option_Type == "Call") %>% pull(Price)
  ) %>%
  ungroup() 

portfolio
```

## One risk driver and uni-variate Gaussian Model
```{r}
#Computing the log-returns
daily_log_returns <- diff(log(sp500))[-1]
 
#Calculate the mean and standard deviation of the daily log-returns
#They serve as estimators for the parameters of the uni-variate Gaussian model
mu_daily_log_returns_asset <- mean(daily_log_returns)
n_asset <- length(daily_log_returns)
sigma_daily_log_returns_asset <- sd(daily_log_returns) * sqrt((n_asset - 1) / n_asset)  # MLE adjustment

# Generate 10,000 scenarios for the underlying asset price in 5 days using the uni-variate Gaussian model
days_ahead <- 5
number_of_scenarios <- 10000

scenarios_asset <- f_normal_scenarios(
  mu_returns = mu_daily_log_returns_asset,
  sigma_returns = sigma_daily_log_returns_asset,
  n_scenarios = number_of_scenarios,
  T = days_ahead,
  spot_price = last_trading_price
)
```


To complement the frequency histograms below, we overlay a kernel density estimate (KDE) on each P&L distribution. The KDE is computed with Silverman's rule-of-thumb bandwidth and places a smooth kernel (Gaussian by default in `geom_density`) at each observation, then sums the contributions. The resulting curve gives a continuous reading of the distribution shape and makes it easier to compare across models. The main caveat is that smoothing is indifferent to the underlying data structure: a wide bandwidth will wash out genuine secondary modes or sharp asymmetries that the raw histogram would reveal. We therefore keep the histogram as the primary visual and treat the density curve as a reading aid, not a replacement.

At t=5, the original interest rates quoted at t=0 are converted to forward rates for accurate option revaluation. We derive the forward curve from the original term structure, which accounts for the time passage and ensures correct discounting of future cash flows.


```{r}

rf_structure_forward <- f_calculate_forward_rates(rf_structure, days_ahead, trading_days_in_year)

# Calculate P&L Distribution using univariate model
portfolio_value_inception <- sum(portfolio$Call_Price)

pnl_results_univariate <- calculate_pnl_univariate(
  scenarios_asset = scenarios_asset,
  portfolio = portfolio,
  rf_structure = rf_structure_forward,
  trading_days_in_year = trading_days_in_year,
  last_volatility = last_volatility,
  portfolio_value_inception = portfolio_value_inception
)

pnl_distribution <- pnl_results_univariate$pnl_distribution
VaR95 <- pnl_results_univariate$VaR95
ES95 <- pnl_results_univariate$ES95

# Plot and save
pnl_distrib_plot <- plot_pnl_univariate(
  pnl_distribution = pnl_distribution,
  VaR95 = VaR95,
  ES95 = ES95,
  output_path = here("Outputs", "pnl_distribution_plot_one_risk_driver.png")
) 


```




## Two risk drivers and bivariate Gaussian model
```{r}
#Compute the log-returns for the volatility index
daily_log_returns_vix <- diff(log(vix))[-1]

#Calculate the mean and standard deviation of the daily log-returns for the volatility index
mu_daily_log_returns_vix <- mean(daily_log_returns_vix)
sigma_daily_log_returns_vix <- sd(daily_log_returns_vix)

# Proper bivariate Gaussian model: simulate (S, VIX) jointly with correlation
rho_gaussian <- cor(as.numeric(daily_log_returns), as.numeric(daily_log_returns_vix))
scenarios_biv <- f_normal_scenarios_bivar(
  mu_returns    = c(mu_daily_log_returns_asset, mu_daily_log_returns_vix),
  sigma_returns = c(sigma_daily_log_returns_asset, sigma_daily_log_returns_vix),
  corr_matrix   = matrix(c(1, rho_gaussian, rho_gaussian, 1), nrow = 2, byrow = TRUE),
  n_scenarios   = number_of_scenarios,
  T            = days_ahead,
  spot_price   = c(last_trading_price, last_volatility)
)

scenarios_asset_biv <- scenarios_biv[, 1]
scenarios_vix       <- scenarios_biv[, 2]


```

```{r}
# Calculate P&L Distribution using bivariate model
pnl_results_bivariate <- calculate_pnl_bivariate(
  scenarios_asset = scenarios_asset_biv,
  scenarios_vix = scenarios_vix,
  portfolio = portfolio,
  rf_structure = rf_structure_forward,
  trading_days_in_year = trading_days_in_year,
  portfolio_value_inception = portfolio_value_inception
)

pnl_distribution_biv <- pnl_results_bivariate$pnl_distribution
VaR95_biv <- pnl_results_bivariate$VaR95
ES95_biv <- pnl_results_bivariate$ES95

# Plot and save
pnl_distrib_plot <- plot_pnl_bivariate(
  pnl_distribution = pnl_distribution_biv,
  VaR95 = VaR95_biv,
  ES95 = ES95_biv,
  output_path = here("Outputs", "pnl_distribution_plot_two_risk_driver.png")
)
```

```{r}
#Sanity check, both PnL distribution are similar but not identical
#Here is the overlaid visualization of the two P&L distributions. For a long call portfolio, you’re delta-positive (hurt when S&P drops) and vega-positive (helped when implied vol/VIX rises). Our fitted dependence is the usual negative correlation between S&P returns and VIX returns, then in bad equity scenarios (SP500 down), we tend to also get VIX up (additionally the VIX would shoot up when SP500 drops due to the leverage effect). This partially offsets losses, so the two-driver VaR/ES is less negative than the univariate “vol fixed” model. 
pnl_combined <- data.frame(
  PnL   = c(pnl_distribution, pnl_distribution_biv),
  Model = rep(c("One driver", "Two drivers"), each = 10000)
)

overlaid_dist <- ggplot(pnl_combined, aes(x = PnL, fill = Model)) +
  geom_histogram(binwidth = 1, alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("One driver" = "#2C3E50", "Two drivers" = "#E74C3C")) +
  theme_minimal()

print(overlaid_dist)

#Save plot in the output folder
ggsave(filename = here("Outputs", "Overlaid_PnL_Distributions.png"), 
       plot = overlaid_dist, width = 8, height = 5, dpi = 300) 
```



##Two risk drivers and copula-marginal model (Student-t and Gaussian copula)
```{r}
# Estimate the correlation between the asset's return and the VIX return and the parameters of the marginal distributions (Student-t in this case) for both the asset and the VIX.
correlation <- cor(daily_log_returns, daily_log_returns_vix)
Corr_matrix <- matrix(data=c(1, correlation, correlation, 1), nrow=2, ncol=2, byrow=TRUE)
sig <- c(1, 1) 
Sigma <- diag(sig) %*% Corr_matrix %*% diag(sig) 

degrees_freedom_asset <- 10 
degrees_freedom_vix <- 5

degrees_freedom_vector <- c(degrees_freedom_asset, degrees_freedom_vix)

#Generate standard bivariate normal draws - 10 000 pairs from a bivariate normal distribution with the estimated correlation
bvn_samples <- mvrnorm(n = number_of_scenarios, mu = c(0,0), Sigma = Sigma)

#Transform to uniform space via pnorm
uniform_samples <- pnorm(bvn_samples)

#Sanity check - visualize the bivariate normal samples and the corresponding uniform samples to confirm that the transformation is working as expected
par(mfrow = c(1, 2))
plot(bvn_samples, pch = 19, col = "blue", main = "Bivariate Normal Distribution")
plot(uniform_samples, pch = 19, col = "red", main = "Correlated Uniform Samples")
par(mfrow = c(1, 1))

#Transform to Student-t space via qt
student_draws <- data.frame(
  Asset = qt(uniform_samples[, 1], df = 10),
  VIX   = qt(uniform_samples[, 2], df = 5)
)

#Scale to real returns
corr_draws <- data.frame(
  Asset = student_draws$Asset / sqrt(10/8) * sigma_daily_log_returns_asset + mu_daily_log_returns_asset,
  VIX   = student_draws$VIX   / sqrt(5/3)  * sigma_daily_log_returns_vix   + mu_daily_log_returns_vix
)
```



```{r}
copula_plot <- ggplot() + theme_void()  

p1 <- ggplot(student_draws, aes(x = Asset, y = VIX)) +
  geom_point(color = rgb(0, 0, 1, 0.3), size = 0.4) +
  labs(title = "Student-t Draws (standardized)",
       x = "Asset (t, df=10)", y = "VIX (t, df=5)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 10))

p2 <- ggplot(student_draws, aes(x = Asset)) +
  geom_histogram(bins = 60, fill = "#2C3E50", color = NA) +
  labs(title = "Asset margin (t, df=10)", x = "", y = "Frequency") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 10))

p3 <- ggplot(student_draws, aes(x = VIX)) +
  geom_histogram(bins = 60, fill = "#E74C3C", color = NA) +
  labs(title = "VIX margin (t, df=5)", x = "", y = "Frequency") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 10))

p4 <- ggplot(corr_draws, aes(x = Asset, y = VIX)) +
  geom_point(color = rgb(0, 0.5, 0, 0.3), size = 0.4) +
  labs(title = "Scaled Draws (real return units)",
       x = "Asset return", y = "VIX return") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 10))

copula_plot <- (p2 + p3) / (p1 + p4) +
  plot_annotation(
    title    = "Gaussian Copula with Student-t Marginals",
    subtitle = "Asset: t(df=10) | VIX: t(df=5) | Normal copula dependence",
    caption  = "Top: marginal distributions | Bottom: joint dependence structure",
    theme = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "#566573", size = 10),
      plot.caption  = element_text(color = "#566573", size = 9, hjust = 0)
    )
  )

print(copula_plot)

ggsave(filename = here("Outputs", "copula_plot.png"), 
       plot = copula_plot , width = 8, height = 5, dpi = 300) 
```

```{r}

scenarios_asset_cop <- last_trading_price * exp(corr_draws$Asset)
scenarios_vix_cop   <- last_volatility * exp(corr_draws$VIX)

#Calculate the option prices at the end of the 5 days for each scenario including vix scenarios from the copula-marginal model
portfolio_value_five_days_ahead_copula <- mapply(function(scenarios_asset_cop, scenario_vix_cop) {
  sum(
    mapply(function(strike, maturity) {
      f_black_scholes(
        S = scenarios_asset_cop,
        K = strike,
        T = (maturity - 5) / trading_days_in_year,
        rf_structure = rf_structure,
        sigma = scenario_vix_cop        
      ) %>% filter(Option_Type == "Call") %>% pull(Price)
    }, portfolio$Strike_Price, portfolio$Maturity_days)
  )
}, scenarios_asset_cop, scenarios_vix_cop)

# Calculate the P&L distribution
pnl_distribution_copula <- portfolio_value_five_days_ahead_copula - portfolio_value_inception

# Compute the VaR95 and ES95
VaR95_copula <- quantile(pnl_distribution_copula, 0.05)
ES95_copula  <- mean(pnl_distribution_copula[pnl_distribution_copula <= VaR95_copula])

# Plot the P&L distribution
pnl_df_copula <- data.frame(PnL = pnl_distribution_copula)

pnl_distrib_plot_copula <- ggplot(pnl_df_copula, aes(x = PnL)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 1, fill = "#2C3E50", color = "#AEB6BF", alpha = 0.85) +
  geom_density(color = "#E74C3C", linewidth = 0.8, adjust = 1) +
  geom_vline(xintercept = VaR95_copula, color = "cyan", linewidth = 0.9) +
  geom_vline(xintercept = ES95_copula,  color = "darkblue", linewidth = 0.9) +
  annotate("text", x = VaR95_copula, y = Inf, 
           label = paste0("VaR 95%: $", round(VaR95_copula, 2)),
           color = "cyan", hjust = -0.1, vjust = 2, size = 3.5, fontface = "bold") +
  annotate("text", x = ES95_copula, y = Inf,  
           label = paste0("ES 95%: $",  round(ES95_copula, 2)),
           color = "darkblue", hjust = -0.1, vjust = 4, size = 3.5, fontface = "bold") +
  labs(title    = "P&L Distribution — Portfolio of European Call Options",
       subtitle = paste0("10,000 scenarios | 5-day horizon | ",
                         "VaR 95%: $", round(VaR95_copula, 2), 
                         " | ES 95%: $", round(ES95_copula, 2)),
       x = "P&L ($)", 
       y = "Density",
       caption = "Copula-marginal model — two risk drivers | Red curve: KDE (Silverman bandwidth)") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "#566573", size = 10),
    plot.caption  = element_text(color = "#566573", size = 9, hjust = 0),
    panel.grid.minor = element_blank()
  )

print(pnl_distrib_plot_copula)

#Save plot in the output folder
ggsave(filename = here("Outputs", "pnl_distribution_plot_copula.png"),
       plot = pnl_distrib_plot_copula, width = 8, height = 5, dpi = 300)
```



##Volatility Surface

Fit a volatility surface to the implied volatilities observed on the market (traded call and put options).
Minimize the absolute distance between the market implied volatilities and the model implied volatilities.
The parametric surface is given by:
      σ(m, τ ) = α1 + α2(m − 1)2 + α3(m − 1)3 + α4√τ ,
      
where m = K/S is the moneyness and τ is the time to maturity of the option in years.
```{r Prepare data for volatility surface fitting, message=FALSE, warning=FALSE}
# Use only OTM calls and OTM puts for surface calibration
options_data <- prepare_options_data_otm(
  calls = calls,
  puts = puts,
  spot_price = last_trading_price
)
```

```{r}
# Fit the parametric surface on OTM quotes
fit_surface <- fit_vol_surface_otm(
  options_data = options_data,
  last_volatility = last_volatility
)

alpha_hat <- fit_surface$alpha_hat
print(alpha_hat)
```

```{r}
# 3D visualization: fitted surface + market IV points (OTM)
iv_surface_3d <- plot_vol_surface_3d(
  options_data = options_data,
  alpha_hat = alpha_hat,
  output_path = here("Outputs", "iv_surface_3d.html")
)

iv_surface_3d
```

Re-price the portfolio using the parametric model. Once you have fitted the volatility surface, consider
˜σ(m, τ ) = σ(m, τ ) − (α1 + α4) + VIXT , where VIXT is the VIX at time T . Use this for the pricing at
time T . For the pricing at T + 5, consider ˜σ(m, τ ) = σ(m, τ ) − (α1 + α4) + VIXT +5 where VIXT +5 is
obtained by simulation.

```{r}
# Re-price the portfolio using the fitted parametric surface sigma_hat(m, tau) (no VIX adjustment yet)
surface_prices_sigma_hat_T <- portfolio_surface_prices_sigma_hat(
  portfolio = portfolio,
  alpha_hat = alpha_hat,
  spot_price = last_trading_price,
  rf_structure = rf_structure,
  trading_days_in_year = trading_days_in_year,
  horizon_days = 0
)

surface_prices_sigma_hat_T$portfolio_value
```


```{r}
# Build adjusted volatility at time T for the portfolio (sigma to plug into BMS)
surface_inputs_T <- portfolio_surface_inputs(
  portfolio = portfolio,
  alpha_hat = alpha_hat,
  spot_price = last_trading_price,
  vix_level = last_volatility,
  trading_days_in_year = trading_days_in_year,
  horizon_days = 0
)

surface_inputs_T
```

```{r}
# Vol surface P&L using bivariate Gaussian scenarios (spot + VIX)
pnl_surface_biv <- calculate_pnl_vol_surface(
  portfolio = portfolio,
  alpha_hat = alpha_hat,
  spot_T = last_trading_price,
  vix_T = last_volatility,
  rf_structure_T = rf_structure,
  rf_structure_T_plus_h = rf_structure_forward,
  trading_days_in_year = trading_days_in_year,
  horizon_days = days_ahead,
  spot_scenarios_T_plus_h = scenarios_asset_biv,
  vix_scenarios_T_plus_h = scenarios_vix
)

pnl_distribution_surface_biv <- pnl_surface_biv$pnl_distribution
VaR95_surface_biv <- pnl_surface_biv$VaR95
ES95_surface_biv  <- pnl_surface_biv$ES95

plot_pnl_vol_surface(
  pnl_distribution = pnl_distribution_surface_biv,
  VaR95 = VaR95_surface_biv,
  ES95 = ES95_surface_biv,
  horizon_days = days_ahead,
  caption = "Vol surface model — VIX from bivariate Gaussian (spot + VIX jointly simulated)",
  output_path = here("Outputs", "pnl_distribution_vol_surface_bivariate.png")
)
```

===========================Annexe=================================

```{r visualize the volatility surface, message=FALSE, warning=FALSE}
# Cleaning filters — kept separate so options_data remains intact
IV_MAX        <- 0.8
MONEYNESS_MIN <- 0.7
MONEYNESS_MAX <- 1.3
CALL_M_MAX    <- 1.15
PUT_M_MIN     <- 0.85

options_data_clean <- options_data %>%
  filter(Implied_Volatility < IV_MAX,
         Moneyness > MONEYNESS_MIN,
         Moneyness < MONEYNESS_MAX)

options_data_calls <- options_data_clean %>% filter(Type == "Call", Moneyness < CALL_M_MAX)
options_data_puts  <- options_data_clean %>% filter(Type == "Put",  Moneyness > PUT_M_MIN)

# Shared fitted surface grid
f_make_surface_grid <- function(data) {
  moneyness_grid <- seq(min(data$Moneyness), max(data$Moneyness), length.out = 50)
  tau_grid       <- seq(min(data$Tau),       max(data$Tau),       length.out = 50)
  surface_grid   <- outer(moneyness_grid, tau_grid, function(m, tau) f_vol_surface(alpha_hat, m, tau))
  list(m = moneyness_grid, tau = tau_grid, z = surface_grid)
}

# Shared plot builder
f_vol_surface_plot <- function(data, title) {
  grid <- f_make_surface_grid(data)
  plot_ly() %>%
    add_surface(
      x = grid$m, y = grid$tau, z = t(grid$z),
      opacity    = 0.6,
      colorscale = list(c(0, 1), c("#2C3E50", "#E74C3C")),
      showscale  = FALSE,
      name       = "Fitted Surface"
    ) %>%
    add_trace(
      data = data,
      x = ~Moneyness, y = ~Tau, z = ~Implied_Volatility,
      type = "scatter3d", mode = "markers",
      marker = list(
        size       = 2,
        color      = ~Implied_Volatility,
        colorscale = list(c(0, 1), c("#AEB6BF", "#E74C3C")),
        showscale  = TRUE,
        colorbar   = list(title = "IV")
      ),
      name = "Market IV"
    ) %>%
    layout(
      title = title,
      scene = list(
        xaxis  = list(title = "Moneyness (K/S)"),
        yaxis  = list(title = "Tau (years)"),
        zaxis  = list(title = "Implied Volatility"),
        camera = list(eye = list(x = 1.8, y = -1.8, z = 0.8))
      )
    )
}

vol_surface_calls <- f_vol_surface_plot(options_data_calls, "Implied Volatility Surface — Calls")
vol_surface_puts  <- f_vol_surface_plot(options_data_puts,  "Implied Volatility Surface — Puts")

vol_surface_calls
vol_surface_puts
```
