---
title: "TP2 - MATH60633A.H2026 - Statistical Methods for Financial Data "
author: "Francis Mazzone () & Vaast Émilien (11364986)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document presents the analysis of the data collected for our project. We will explore the data, perform statistical tests, and visualize the results to draw meaningful conclusions.

```{r libraries, message=FALSE, echo= TRUE}
# Load necessary libraries
library(quantmod)
library(knitr)
#install.packages("patchwork")
library(patchwork)
#install.packages("kableExtra")
library(kableExtra)
#install.packages("here")
library(here)
library(xts)
#install.packages("PerformanceAnalytics")
library(PerformanceAnalytics)
library(zoo)
#install.packages("dplyr")
#install.packages("lubridate")
library(dplyr)
library(lubridate)
library(ggplot2)
library(devtools)
library(tidyverse)
#install.packages("MASS")
library(MASS)
```

```{r environment set up, message=FALSE, warning=FALSE}
data_env <- new.env()
source(here("Functions", "f_black_scholes.R"))
source(here("Functions", "f_normal_scenarios.R"))

file_path <- here("Data", "Raw", "Market.rda")
stopifnot(file.exists(file_path))

load(file_path, envir = data_env)
ls(data_env)

data <- data_env$Market

sp500 <- data$sp500
vix <- data$vix 

rf_structure <- data.frame(
  y_maturity = as.numeric(attr(data[["rf"]], "names")),
  rate = as.numeric(data[["rf"]])
)

calls <- data$calls
puts <- data$puts

```

## Pricing of a portfolio of options

```{r define conventions, message=FALSE, warning=FALSE}
# Define conventions
days_in_year <- 360
trading_days_in_year <- 250
```

We want to build a portfolio that consists of 4 European call options:

-   Strike price: 1600 \| Maturity: 20 days
-   Strike price: 1650 \| Maturity: 20 days
-   Strike price: 1750 \| Maturity: 40 days
-   Strike price: 1800 \| Maturity: 40 days

```{r Build the portfolio, message=FALSE, warning=FALSE}
#Underlying asset price and volatility for the options in the portfolio
last_trading_price <- coredata(tail(sp500, 1))
last_volatility <- coredata(tail(vix, 1))

# Define the portfolio with the specified options
maturities_day = c(20,20,40,40) 
strike_prices = c(1600, 1650, 1750, 1800)

#Build the portfolio data frame
portfolio <- data.frame(
  Maturity_days = maturities_day, 
  Strike_Price = strike_prices,
  Trading_Price = rep(last_trading_price, 4),
  Volatility = rep(last_volatility, 4)
)
```

```{r Calculate option prices, message=FALSE, warning=FALSE}
# Calculate the option prices for each option in the portfolio using the Black-Scholes formula
portfolio <- portfolio %>% 
  rowwise() %>%
  mutate(
    Call_Price = f_black_scholes(
      S    = Trading_Price,
      K    = Strike_Price,
      T    = Maturity_days / trading_days_in_year,
      trading_day_convention = trading_days_in_year,
      rate_day_convention = days_in_year,
      rf_structure = rf_structure,
      sigma = Volatility
    ) %>% filter(Option_Type == "Call") %>% pull(Price)
  ) %>%
  ungroup() 

portfolio
```

## One risk driver and uni-variate Gaussian Model
```{r}
#Computing the log-returns
daily_log_returns <- diff(log(sp500))[-1]
 
#Calculate the mean and standard deviation of the daily log-returns
#They serve as estimators for the parameters of the uni-variate Gaussian model
mu_daily_log_returns_asset <- mean(daily_log_returns)
sigma_daily_log_returns_asset <- sd(daily_log_returns)

# Generate 10,000 scenarios for the underlying asset price in 5 days using the uni-variate Gaussian model
days_ahead <- 5
number_of_scenarios <- 10000

scenarios_asset <- f_normal_scenarios(
  mu_returns = mu_daily_log_returns_asset,
  sigma_returns = sigma_daily_log_returns_asset,
  n_scenarios = number_of_scenarios,
  T = days_ahead,
  spot_price = last_trading_price
)
```


```{r}
#Determine the P&L distribution of the book of options, using the simulated underlying values. Assume
#the implied volatility stays the same. Take interpolated rates for the term structure.

portfolio_value_inception <- sum(portfolio$Call_Price)
# Calculate the option prices at the end of the 5 days for each scenario
portfolio_value_five_days_ahead <- sapply(scenarios_asset, function(scenario_price) {
  sum(
    mapply(function(strike, maturity) {
      f_black_scholes(
        S = scenario_price,
        K = strike,
        T = (maturity - 5) / trading_days_in_year,
        rf_structure = rf_structure,
        sigma = last_volatility
      ) %>% filter(Option_Type == "Call") %>% pull(Price)
    }, portfolio$Strike_Price, portfolio$Maturity_days)
  )
})

# Calculate the P&L distribution
pnl_distribution <- portfolio_value_five_days_ahead - portfolio_value_inception

# Compute the VaR95 and ES95
VaR95 <- quantile(pnl_distribution, 0.05)
ES95  <- mean(pnl_distribution[pnl_distribution <= VaR95])

# Plot the P&L distribution
pnl_df <- data.frame(PnL = pnl_distribution)

pnl_distrib_plot <- ggplot(pnl_df, aes(x = PnL)) +
  geom_histogram(binwidth = 1, fill = "#2C3E50", color = "#AEB6BF", alpha = 0.85) +
  geom_vline(xintercept = VaR95, color = "cyan", linewidth = 0.9) +
  geom_vline(xintercept = ES95,  color = "darkblue", linewidth = 0.9) +
  annotate("text", x = VaR95, y = Inf, 
           label = paste0("VaR 95%: $", round(VaR95, 2)),
           color = "cyan", hjust = -0.1, vjust = 2, size = 3.5, fontface = "bold") +
  annotate("text", x = ES95, y = Inf,  
           label = paste0("ES 95%: $",  round(ES95, 2)),
           color = "darkblue", hjust = -0.1, vjust = 4, size = 3.5, fontface = "bold") +
  labs(title    = "P&L Distribution — Portfolio of European Call Options",
       subtitle = paste0("10,000 scenarios | 5-day horizon | ",
                         "VaR 95%: $", round(VaR95, 2), 
                         " | ES 95%: $", round(ES95, 2)),
       x = "P&L ($)", 
       y = "Frequency",
       caption = "Univariate Gaussian model — one risk driver") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "#566573", size = 10),
    plot.caption  = element_text(color = "#566573", size = 9, hjust = 0),
    panel.grid.minor = element_blank()
  )

print(pnl_distrib_plot)

#Save plot in the output folder
ggsave(filename = here("Outputs", "pnl_distribution_plot_one_risk_driver.png"), 
       plot = pnl_distrib_plot, width = 8, height = 5, dpi = 300) 


```


## Two risk drivers and bivariate Gaussian model
```{r}
#Compute the log-returns for the volatility index
daily_log_returns_vix <- diff(log(vix))[-1]

#Calculate the mean and standard deviation of the daily log-returns for the volatility index
mu_daily_log_returns_vix <- mean(daily_log_returns_vix)
sigma_daily_log_returns_vix <- sd(daily_log_returns_vix)

scenarios_vix <- f_normal_scenarios(
  mu_returns = mu_daily_log_returns_vix,
  sigma_returns = sigma_daily_log_returns_vix,
  n_scenarios = number_of_scenarios,
  T = days_ahead,
  spot_price = last_volatility
)


```

```{r}
# Calculate the option prices at the end of the 5 days for each scenario including vix scenarios
portfolio_value_five_days_ahead_vix <- mapply(function(scenario_price, scenario_vix) {
  sum(
    mapply(function(strike, maturity) {
      f_black_scholes(
        S = scenario_price,
        K = strike,
        T = (maturity - 5) / trading_days_in_year,
        rf_structure = rf_structure,
        sigma = scenario_vix        
      ) %>% filter(Option_Type == "Call") %>% pull(Price)
    }, portfolio$Strike_Price, portfolio$Maturity_days)
  )
}, scenarios_asset, scenarios_vix)

# Calculate the P&L distribution
pnl_distribution_biv <- portfolio_value_five_days_ahead_vix - portfolio_value_inception

# Compute the VaR95 and ES95
VaR95_biv <- quantile(pnl_distribution_biv, 0.05)
ES95_biv  <- mean(pnl_distribution_biv[pnl_distribution_biv <= VaR95])

# Plot the P&L distribution
pnl_df_biv <- data.frame(PnL = pnl_distribution_biv)

pnl_distrib_plot <- ggplot(pnl_df_biv, aes(x = PnL)) +
  geom_histogram(binwidth = 1, fill = "#2C3E50", color = "#AEB6BF", alpha = 0.85) +
  geom_vline(xintercept = VaR95_biv, color = "cyan", linewidth = 0.9) +
  geom_vline(xintercept = ES95_biv,  color = "darkblue", linewidth = 0.9) +
  annotate("text", x = VaR95_biv, y = Inf, 
           label = paste0("VaR 95%: $", round(VaR95_biv, 2)),
           color = "cyan", hjust = -0.1, vjust = 2, size = 3.5, fontface = "bold") +
  annotate("text", x = ES95_biv, y = Inf,  
           label = paste0("ES 95%: $",  round(ES95_biv, 2)),
           color = "darkblue", hjust = -0.1, vjust = 4, size = 3.5, fontface = "bold") +
  labs(title    = "P&L Distribution — Portfolio of European Call Options",
       subtitle = paste0("10,000 scenarios | 5-day horizon | ",
                         "VaR 95%: $", round(VaR95_biv, 2), 
                         " | ES 95%: $", round(ES95_biv, 2)),
       x = "P&L ($)", 
       y = "Frequency",
       caption = "Univariate Gaussian model — two risk driver") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "#566573", size = 10),
    plot.caption  = element_text(color = "#566573", size = 9, hjust = 0),
    panel.grid.minor = element_blank()
  )

print(pnl_distrib_plot)

#Save plot in the output folder
ggsave(filename = here("Outputs", "pnl_distribution_plot_two_risk_driver.png"), 
       plot = pnl_distrib_plot, width = 8, height = 5, dpi = 300) 
```

```{r}
#Sanity check, both PnL distribution are very similar but not identical
#Here is the overlaid visualization of the two P&L distributions, we can see that the bivariate model has a slightly wider distribution, which is consistent with the fact that it incorporates more risk factors.
pnl_combined <- data.frame(
  PnL   = c(pnl_distribution, pnl_distribution_biv),
  Model = rep(c("One driver", "Two drivers"), each = 10000)
)

overlaid_dist <- ggplot(pnl_combined, aes(x = PnL, fill = Model)) +
  geom_histogram(binwidth = 1, alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("One driver" = "#2C3E50", "Two drivers" = "#E74C3C")) +
  theme_minimal()

print(overlaid_dist)

#Save plot in the output folder
ggsave(filename = here("Outputs", "Overlaid_PnL_Distributions.png"), 
       plot = overlaid_dist, width = 8, height = 5, dpi = 300) 
```



##Two risk drivers and copula-marginal model (Student-t and Gaussian copula)
```{r}
# Estimate the correlation between the asset's return and the VIX return and the parameters of the marginal distributions (Student-t in this case) for both the asset and the VIX.
correlation <- cor(daily_log_returns, daily_log_returns_vix)
Corr_matrix <- matrix(data=c(1, correlation, correlation, 1), nrow=2, ncol=2, byrow=TRUE)
sig <- c(1, 1) 
Sigma <- diag(sig) %*% Corr_matrix %*% diag(sig) 

degrees_freedom_asset <- 10 
degrees_freedom_vix <- 5

degrees_freedom_vector <- c(degrees_freedom_asset, degrees_freedom_vix)

#Generate standard bivariate normal draws - 10 000 pairs from a bivariate normal distribution with the estimated correlation
bvn_samples <- mvrnorm(n = number_of_scenarios, mu = c(0,0), Sigma = Sigma)

#Transform to uniform space via pnorm
uniform_samples <- pnorm(bvn_samples)

#Sanity check - visualize the bivariate normal samples and the corresponding uniform samples to confirm that the transformation is working as expected
par(mfrow = c(1, 2))
plot(bvn_samples, pch = 19, col = "blue", main = "Bivariate Normal Distribution")
plot(uniform_samples, pch = 19, col = "red", main = "Correlated Uniform Samples")
par(mfrow = c(1, 1))

#Transform to Student-t space via qt
student_draws <- data.frame(
  Asset = qt(uniform_samples[, 1], df = 10),
  VIX   = qt(uniform_samples[, 2], df = 5)
)

#Scale to real returns
corr_draws <- data.frame(
  Asset = student_draws$Asset / sqrt(10/8) * sigma_daily_log_returns_asset + mu_daily_log_returns_asset,
  VIX   = student_draws$VIX   / sqrt(5/3)  * sigma_daily_log_returns_vix   + mu_daily_log_returns_vix
)
```



```{r}
copula_plot <- ggplot() + theme_void()  

p1 <- ggplot(student_draws, aes(x = Asset, y = VIX)) +
  geom_point(color = rgb(0, 0, 1, 0.3), size = 0.4) +
  labs(title = "Student-t Draws (standardized)",
       x = "Asset (t, df=10)", y = "VIX (t, df=5)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 10))

p2 <- ggplot(student_draws, aes(x = Asset)) +
  geom_histogram(bins = 60, fill = "#2C3E50", color = NA) +
  labs(title = "Asset margin (t, df=10)", x = "", y = "Frequency") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 10))

p3 <- ggplot(student_draws, aes(x = VIX)) +
  geom_histogram(bins = 60, fill = "#E74C3C", color = NA) +
  labs(title = "VIX margin (t, df=5)", x = "", y = "Frequency") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 10))

p4 <- ggplot(corr_draws, aes(x = Asset, y = VIX)) +
  geom_point(color = rgb(0, 0.5, 0, 0.3), size = 0.4) +
  labs(title = "Scaled Draws (real return units)",
       x = "Asset return", y = "VIX return") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 10))

copula_plot <- (p2 + p3) / (p1 + p4) +
  plot_annotation(
    title    = "Gaussian Copula with Student-t Marginals",
    subtitle = "Asset: t(df=10) | VIX: t(df=5) | Normal copula dependence",
    caption  = "Top: marginal distributions | Bottom: joint dependence structure",
    theme = theme(
      plot.title    = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = "#566573", size = 10),
      plot.caption  = element_text(color = "#566573", size = 9, hjust = 0)
    )
  )

print(copula_plot)

ggsave(filename = here("Outputs", "copula_plot.png"), 
       plot = copula_plot , width = 8, height = 5, dpi = 300) 
```

```{r}

scenarios_asset_cop <- last_trading_price * exp(corr_draws$Asset)
scenarios_vix_cop   <- last_volatility * exp(corr_draws$VIX)

#Calculate the option prices at the end of the 5 days for each scenario including vix scenarios from the copula-marginal model
portfolio_value_five_days_ahead_copula <- mapply(function(scenarios_asset_cop, scenario_vix_cop) {
  sum(
    mapply(function(strike, maturity) {
      f_black_scholes(
        S = scenarios_asset_cop,
        K = strike,
        T = (maturity - 5) / trading_days_in_year,
        rf_structure = rf_structure,
        sigma = scenario_vix_cop        
      ) %>% filter(Option_Type == "Call") %>% pull(Price)
    }, portfolio$Strike_Price, portfolio$Maturity_days)
  )
}, scenarios_asset_cop, scenarios_vix_cop)

# Calculate the P&L distribution
pnl_distribution_copula <- portfolio_value_five_days_ahead_copula - portfolio_value_inception

# Compute the VaR95 and ES95
VaR95_copula <- quantile(pnl_distribution_copula, 0.05)
ES95_copula  <- mean(pnl_distribution_copula[pnl_distribution_copula <= VaR95_copula])

# Plot the P&L distribution
pnl_df_copula <- data.frame(PnL = pnl_distribution_copula)

pnl_distrib_plot_copula <- ggplot(pnl_df_copula, aes(x = PnL)) +
  geom_histogram(binwidth = 1, fill = "#2C3E50", color = "#AEB6BF", alpha = 0.85) +
  geom_vline(xintercept = VaR95_copula, color = "cyan", linewidth = 0.9) +
  geom_vline(xintercept = ES95_copula,  color = "darkblue", linewidth = 0.9) +
  annotate("text", x = VaR95_copula, y = Inf, 
           label = paste0("VaR 95%: $", round(VaR95_copula, 2)),
           color = "cyan", hjust = -0.1, vjust = 2, size = 3.5, fontface = "bold") +
  annotate("text", x = ES95_copula, y = Inf,  
           label = paste0("ES 95%: $",  round(ES95_copula, 2)),
           color = "darkblue", hjust = -0.1, vjust = 4, size = 3.5, fontface = "bold") +
  labs(title    = "P&L Distribution — Portfolio of European Call Options",
       subtitle = paste0("10,000 scenarios | 5-day horizon | ",
                         "VaR 95%: $", round(VaR95_copula, 2), 
                         " | ES 95%: $", round(ES95_copula, 2)),
       x = "P&L ($)", 
       y = "Frequency",
       caption = "Copula-marginal model — two risk drivers") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "#566573", size = 10),
    plot.caption  = element_text(color = "#566573", size = 9, hjust = 0),
    panel.grid.minor = element_blank()
  )

print(pnl_distrib_plot_copula)

#Save plot in the output folder
ggsave(filename = here("Outputs", "pnl_distribution_plot_copula.png"),
       plot = pnl_distrib_plot_copula, width = 8, height = 5, dpi = 300)
```



##Volatility Surface
Fit a volatility surface to the implied volatilities observed on the market (traded call and put options).
Minimize the absolute distance between the market implied volatilities and the model implied volatilities.
The parametric surface is given by:
      σ(m, τ ) = α1 + α2(m − 1)2 + α3(m − 1)3 + α4√τ ,
      
where m = K/S is the moneyness and τ is the time to maturity of the option in years.
```{r}

```

