---
title: "TP2 - MATH60633A.H2026 - Statistical Methods for Financial Data "
author: "Francis Mazzone () & Vaast Émilien (11364986)"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document presents the analysis of the data collected for our project. We will explore the data, perform statistical tests, and visualize the results to draw meaningful conclusions.

```{r libraries, message=FALSE, echo= TRUE}
# Load necessary libraries
library(quantmod)
library(knitr)
#install.packages("patchwork")
library(patchwork)
#install.packages("kableExtra")
library(kableExtra)
#install.packages("here")
library(here)
library(xts)
#install.packages("PerformanceAnalytics")
library(PerformanceAnalytics)
library(zoo)
#install.packages("dplyr")
#install.packages("lubridate")
library(dplyr)
library(lubridate)
library(ggplot2)
library(devtools)
library(tidyverse)
```

```{r environment set up, message=FALSE, warning=FALSE}
data_env <- new.env()
source(here("Functions", "f_black_scholes.R"))

file_path <- here("Data", "Raw", "Market.rda")
stopifnot(file.exists(file_path))

load(file_path, envir = data_env)
ls(data_env)
(data <- data_env$Market)

sp500 <- data$sp500
vix <- data$vix 

rf_structure <- data.frame(
  y_maturity = as.numeric(attr(data[["rf"]], "names")),
  rate = as.numeric(data[["rf"]])
)

calls <- data$calls
puts <- data$puts

```

## Pricing of a portfolio of options

```{r define conventions, message=FALSE, warning=FALSE}
# Define conventions
days_in_year <- 360
trading_days_in_year <- 250
```

We want to build a portfolio that consists of 4 European call options:

-   Strike price: 1600 \| Maturity: 20 days
-   Strike price: 1650 \| Maturity: 20 days
-   Strike price: 1750 \| Maturity: 40 days
-   Strike price: 1800 \| Maturity: 40 days

```{r Build the portfolio, message=FALSE, warning=FALSE}
#Underlying asset price and volatility for the options in the portfolio
last_trading_price <- tail(sp500, 1)
last_volatility <- tail(vix, 1) / 100 

# Define the portfolio with the specified options
maturities_day = c(20,20,40,40) 
strike_prices = c(1600, 1650, 1750, 1800)

#Build the portfolio data frame
portfolio <- data.frame(
  Maturity_days = maturities, 
  Strike_Price = strike_prices,
  Trading_Price = rep(last_trading_price, 4),
  Volatility = rep(last_volatility, 4)
)
```

```{r Calculate option prices, message=FALSE, warning=FALSE}
# Calculate the option prices for each option in the portfolio using the Black-Scholes formula
portfolio <- portfolio %>%
  rowwise() %>%
  mutate(
    Call_Price = f_black_scholes(
      S    = Trading_Price,
      K    = Strike_Price,
      T    = Maturity / trading_days_in_year,
      rf_structure = rf_structure,
      sigma = Volatility
    ) %>% filter(Option_Type == "Call") %>% pull(Price)
  ) %>%
  ungroup() 

portfolio
```

## One risk driver and uni-variate Gaussian Model
