---
title: "TP2 - MATH60633A.H2026 - Statistical Methods for Financial Data "
author: 

  - "Navid Namazi Baigi"
  - "Venkat Narayana Ramakrishnan"
  
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```


```{r libraries, message=FALSE, echo= TRUE}
# Load necessary libraries
library(knitr)
library(patchwork)
library(here)
library(xts)
library(dplyr)
library(purrr)
library(ggplot2)
library(MASS)
library(plotly)
library(mvtnorm)
library(copula)
```

```{r environment set up, message=FALSE, warning=FALSE}
data_env <- new.env()
source(here("Functions", "f_black_scholes.R"))
source(here("Functions", "f_pnl_plot.R"))
source(here("Functions", "f_vol_surface_workflow.R"))
source(here("Functions", "f_copula_diagnostics_plot.R"))

file_path <- here("Data", "Raw", "Market.rda")
stopifnot(file.exists(file_path))

load(file_path, envir = data_env)
# ls(data_env)

data <- data_env$Market

sp500 <- data$sp500
vix <- data$vix 

rf_structure <- data.frame(
  y_maturity = as.numeric(attr(data[["rf"]], "names")),
  rate = as.numeric(data[["rf"]])
)

calls <- data$calls
puts <- data$puts

# check class of all 
data_classes <- sapply(data, class)
knitr::kable(data_classes)

```

## Pricing of a portfolio of options

We want to build a portfolio that consists of 4 European call options:

-   Strike price: 1600 \| Maturity: 20 days
-   Strike price: 1650 \| Maturity: 20 days
-   Strike price: 1750 \| Maturity: 40 days
-   Strike price: 1800 \| Maturity: 40 days

```{r Build the portfolio, message=FALSE, warning=FALSE}
# Define conventions
days_in_year <- 360
trading_days_in_year <- 250

#Underlying asset price and volatility for the options in the portfolio
last_trading_price <- as.numeric(coredata(tail(sp500, 1)))
last_volatility <- as.numeric(coredata(tail(vix, 1)))

# Define the portfolio with the specified options
maturities_day = c(20,20,40,40) 
strike_prices = c(1600, 1650, 1750, 1800)

#Build the portfolio data frame
portfolio <- data.frame(
  Maturity_days = maturities_day, 
  Strike_Price = strike_prices,
  Trading_Price = rep(last_trading_price, 4),
  Volatility_last_vix = rep(last_volatility, 4)
)
```

```{r Calculate option prices, message=FALSE, warning=FALSE}
# Calculate the option prices for each option in the portfolio using the Black-Scholes formula
portfolio <- portfolio %>% 
  rowwise() %>%
  mutate(
    Call_Price = f_black_scholes(
      S    = Trading_Price,
      K    = Strike_Price,
      T    = Maturity_days / trading_days_in_year,
      trading_day_convention = trading_days_in_year,
      rate_day_convention = days_in_year,
      rf_structure = rf_structure,
      sigma = Volatility_last_vix
    ) %>% filter(Option_Type == "Call") %>% pull(Price)
  ) %>%
  ungroup() 

knitr::kable(portfolio)
```

## One risk driver and uni-variate Gaussian Model

### Methodology: MLE Estimation and Time Scaling (same logic used for the bivariate model)

**Parameter Estimation:** Using closed-form maximum likelihood estimators:

- **Mean**: $\hat{\mu} = \frac{1}{n}\sum_{i=1}^{n} R_i$ (=sample mean)

- **Variance (MLE)**: $\hat{\sigma}^2 = \frac{1}{n}\sum_{i=1}^{n} (R_i - \hat{\mu})^2$ ( = sample variance $\times \frac{n-1}{n}$)


**Time-Scaling for Horizon T:** For independent daily returns, the cumulative T-day log-return follows:
$$\sum_{t=1}^{T} R_t \sim N(T \cdot \mu, T \cdot \sigma^2)$$

Therefore, we generate scenarios from $N(T \cdot \hat{\mu}, T \cdot \hat{\sigma})$ and convert to terminal prices via $S_T = S_0 \cdot e^{\text{cum_log_return}}$.

**Note:** At t=5, the original interest rates quoted at t=0 are converted to forward rates for accurate option revaluation. We derive the forward curve from the original term structure, which accounts for the time passage and ensures correct discounting of future cash flows.

```{r}
#Computing the log-returns
daily_log_returns <- diff(log(sp500))[-1]
 
#Calculate MLE estimates of mean and standard deviation of the daily log-returns
mu_daily_log_returns_asset <- mean(daily_log_returns)
n_asset <- length(daily_log_returns)
sigma_daily_log_returns_asset <- sd(daily_log_returns) * sqrt((n_asset - 1) / n_asset)  # MLE adjustment

# Generate 10,000 scenarios for the underlying asset price in 5 days
days_ahead <- 5
number_of_scenarios <- 10000

# Generate T-day cumulative log-returns from N(T*mu, T*sigma^2)
cum_log_returns <- rnorm(number_of_scenarios, 
                         mean = mu_daily_log_returns_asset * days_ahead,
                         sd = sigma_daily_log_returns_asset * sqrt(days_ahead))

# Convert to prices
scenarios_asset <- last_trading_price * exp(cum_log_returns)

# Calculate forward rates for accurate pricing at T+5
rf_structure_forward <- f_calculate_forward_rates(rf_structure, days_ahead, trading_days_in_year)

knitr::kable(data.frame(
  Parameter = c("Mean (daily)", "Std Dev (daily)"),
  Estimate = c(round(mu_daily_log_returns_asset, 6), round(sigma_daily_log_returns_asset, 6))
), caption = "MLE Estimates for Daily Log-Returns of the Underlying Asset:")
```


To complement the frequency histograms below, we overlay a kernel density estimate (KDE) on each P&L distribution. The KDE is computed with Silverman's rule-of-thumb bandwidth and places a smooth kernel (Gaussian by default in `geom_density`) at each observation, then sums the contributions. The resulting curve gives a continuous reading of the distribution shape and makes it easier to compare across models. The main caveat is that smoothing is indifferent to the underlying data structure: a wide bandwidth will wash out genuine secondary modes or sharp asymmetries that the raw histogram would reveal. We therefore keep the histogram as the primary visual and treat the density curve as a reading aid, not a replacement.


```{r}


# Calculate P&L Distribution using univariate model
portfolio_value_inception <- sum(portfolio$Call_Price)

pnl_results_univariate <- calculate_pnl_univariate(
  scenarios_asset = scenarios_asset,
  portfolio = portfolio,
  rf_structure = rf_structure_forward,
  trading_days_in_year = trading_days_in_year,
  last_volatility = last_volatility,
  portfolio_value_inception = portfolio_value_inception
)

pnl_distribution <- pnl_results_univariate$pnl_distribution
VaR95 <- pnl_results_univariate$VaR95
ES95 <- pnl_results_univariate$ES95

# Plot and save
pnl_distrib_plot <- plot_pnl_univariate(
  pnl_distribution = pnl_distribution,
  VaR95 = VaR95,
  ES95 = ES95,
  output_path = here("Outputs", "pnl_distribution_plot_one_risk_driver.png")
) 


```




## Two risk drivers and bivariate Gaussian model
```{r}

# Compute log-returns for the volatility index
daily_log_returns_vix <- diff(log(vix))[-1]

# Estimate parameters from bivariate returns
combined_returns <- cbind(daily_log_returns, daily_log_returns_vix)
n_obs <- nrow(combined_returns)
mu_mnorm <- colMeans(combined_returns)

# MLE estimate of covariance (divide by n instead of n-1)
cov_mle <- cov(combined_returns) * (n_obs - 1) / n_obs

# Scale to T-day horizon: mu_T = T*mu_daily, Sigma_T = T*Sigma_daily
sigma_mvtnorm <- cov_mle * days_ahead
mu_mnorm_scaled <- mu_mnorm * days_ahead

# Generate scenarios using mvtnorm::rmvnorm
scenarios_mnorm <- mvtnorm::rmvnorm(n = number_of_scenarios, mean = mu_mnorm * days_ahead, sigma = cov_mle * days_ahead)

# Convert to prices
scenarios_asset_biv <- last_trading_price * exp(scenarios_mnorm[, 1])
scenarios_vix <- last_volatility * exp(scenarios_mnorm[, 2])

#kable parametersestimated
knitr::kable(data.frame(
  Parameter = c("Asset Mean (daily)", "Asset Std Dev (daily)", "VIX Mean (daily)", "VIX Std Dev (daily)", "Covariance", "Correlation"),
  Estimate = c(
    round(mu_mnorm[1], 6), 
    round(sqrt(cov_mle[1, 1]), 6), 
    round(mu_mnorm[2], 6), 
    round(sqrt(cov_mle[2, 2]), 6), 
    round(cov_mle[1, 2], 6), 
    round(cov_mle[1, 2] / (sqrt(cov_mle[1, 1]) * sqrt(cov_mle[2, 2])), 6)
  )
), caption = "MLE Estimates for Bivariate Gaussian Model:")
```

```{r}
# Calculate P&L Distribution using bivariate model
pnl_results_bivariate <- calculate_pnl_bivariate(
  scenarios_asset = scenarios_asset_biv,
  scenarios_vix = scenarios_vix,
  portfolio = portfolio,
  rf_structure = rf_structure_forward,
  trading_days_in_year = trading_days_in_year,
  portfolio_value_inception = portfolio_value_inception
)

pnl_distribution_biv <- pnl_results_bivariate$pnl_distribution
VaR95_biv <- pnl_results_bivariate$VaR95
ES95_biv <- pnl_results_bivariate$ES95

# Plot and save
pnl_distrib_plot <- plot_pnl_bivariate(
  pnl_distribution = pnl_distribution_biv,
  VaR95 = VaR95_biv,
  ES95 = ES95_biv,
  output_path = here("Outputs", "pnl_distribution_plot_two_risk_driver.png")
)
```

## Comparison of Univariate vs Bivariate Models: Impact of Volatility Correlation

The bivariate model captures a key market dynamic: the negative correlation between equity returns and implied volatility. This creates a natural hedging benefit for long call portfolios:

- **Delta Effect**: Long calls are delta-positive → losses occur when S&P 500 declines
- **Vega Effect**: Long calls are vega-positive → gains occur when volatility increases
- **Market Structure**: During equity downturns, volatility (VIX) typically spikes (sp500 returns are negatively correlated with vix returns) 
- **Result**: When adverse equity scenarios occur (S&P down), the simultaneous VIX increase partially offsets the losses through vega gains

This natural hedging mechanism causes the bivariate model's risk metrics (VaR, ES) to be less negative (higher value) compared to the univariate model, where volatility is held artificially constant.


## Two risk drivers and copula-marginal model (Student-t and Gaussian copula)

### Step 1: Fit Student-t marginal distributions to each series separately
```{r}
df_asset <- 10
df_vix <- 5

# Fit Student-t distribution to Asset returns using maximum likelihood estimation
# Using MASS::fitdistr to estimate the parameters
fit_asset <- suppressWarnings( fitdistr(daily_log_returns, densfun = "t", df = df_asset) )
mu_asset <- fit_asset$estimate[["m"]]
scale_asset <- fit_asset$estimate[["s"]]
sigma_asset <- scale_asset * sqrt(df_asset / (df_asset - 2))  # Convert scale to standard deviation

# Fit Student-t distribution to VIX returns
fit_vix <- suppressWarnings( fitdistr(daily_log_returns_vix, densfun = "t", df = df_vix) )
mu_vix <- fit_vix$estimate[["m"]]
scale_vix <- fit_vix$estimate[["s"]]
sigma_vix <- scale_vix * sqrt(df_vix / (df_vix - 2))  # Convert scale to standard deviation

# Print estimated parameters 
knitr::kable(data.frame(
  Parameter = c("Asset", "VIX"),
  df = c(df_asset, df_vix),
  mu = c(round(mu_asset, 6), round(mu_vix, 6)),
  scale = c(round(scale_asset, 6), round(scale_vix, 6)),
  sigma = c(round(sigma_asset, 6), round(sigma_vix, 6))
), caption = "Estimated Parameters for Student-t Marginals:")


```

### Step 2: Transform each series to uniform [0,1] space via thier CDFs and Fit Gaussian Copula
```{r}
# Apply Student-t CDF with estimated parameters to transform to uniform samples
# pt() expects standardized input: (x - location) / scale
U_asset <- pt((daily_log_returns - mu_asset) / scale_asset, df = df_asset)

U_vix <- pt((daily_log_returns_vix - mu_vix) / scale_vix, df = df_vix)

# Combined uniform samples
uniforms <- data.frame(
  U_Asset = U_asset,
  U_VIX = U_vix
)
# Combine uniform samples into matrix
U <- cbind(uniforms$U_Asset, uniforms$U_VIX)

# Fit Gaussian copula using maximum likelihood
C <- copula::normalCopula(dim = 2)
fit_copula <- copula::fitCopula(C, data = U, method = "ml")

# Extract estimated correlation parameter
copula_correlation <- coef(fit_copula)
cat("Estimated Gaussian Copula Correlation: ", round(copula_correlation, 6), "\n")
```


### Step 3: Simulate from copula-marginal model for 5-day horizon
```{r}
# Generate 5 days of returns for each scenario from the fitted Gaussian copula
# We need 5 * number_of_scenarios
U_sim_total <- copula::rCopula(n = 5 * number_of_scenarios, copula = fit_copula@copula)

# Transform back to Student-t marginals
rets_asset_daily <- mu_asset + scale_asset * qt(U_sim_total[, 1], df = df_asset)
rets_vix_daily <- mu_vix + scale_vix * qt(U_sim_total[, 2], df = df_vix)

# Reshape to get 5 days
rets_asset_daily <- matrix(rets_asset_daily, nrow = number_of_scenarios, ncol = 5, byrow = TRUE)
rets_vix_daily <- matrix(rets_vix_daily, nrow = number_of_scenarios, ncol = 5, byrow = TRUE)

# Sum daily returns across 5 days to get cumulative log-returns
cum_rets_asset <- rowSums(rets_asset_daily)
cum_rets_vix <- rowSums(rets_vix_daily)

# Convert cumulative log-returns to prices at T+5
scenarios_asset_cop <- last_trading_price * exp(cum_rets_asset)
scenarios_vix_cop <- last_volatility * exp(cum_rets_vix)


# Calculate P&L Distribution using copula-marginal model
pnl_results_copula <- calculate_pnl_bivariate(
  scenarios_asset = scenarios_asset_cop,
  scenarios_vix = scenarios_vix_cop,
  portfolio = portfolio,
  rf_structure = rf_structure_forward,
  trading_days_in_year = trading_days_in_year,
  portfolio_value_inception = portfolio_value_inception
)

pnl_distribution_cop <- pnl_results_copula$pnl_distribution
VaR95_cop <- pnl_results_copula$VaR95
ES95_cop <- pnl_results_copula$ES95

# Plot and save
pnl_distrib_plot <- plot_pnl_copula(
  pnl_distribution = pnl_distribution_cop,
  VaR95 = VaR95_cop,
  ES95 = ES95_cop,
  output_path = here("Outputs", "pnl_distribution_plot_copula_marginal.png")
)

```

---

### Plots: Joint distribution, joint uniforms(pseudo data)- not required in the assignment (just for visualization and sanity check of the copula fit)
```{r, fig.width=14, fig.height=12}
copula_diag_plot <- plot_copula_fit_diagnostics(
  daily_log_returns = daily_log_returns,
  daily_log_returns_vix = daily_log_returns_vix,
  df_asset = df_asset,
  df_vix = df_vix,
  mu_asset = mu_asset,
  mu_vix = mu_vix,
  scale_asset = scale_asset,
  scale_vix = scale_vix,
  sigma_asset = sigma_asset,
  sigma_vix = sigma_vix,
  uniforms = uniforms,
  copula_correlation = copula_correlation,
  output_path = here("Outputs", "copula_plot.png")
)
```


## Volatility Surface

Fit a volatility surface to the implied volatilities observed on the market (traded call and put options).
Minimize the absolute distance between the market implied volatilities and the model implied volatilities.
The parametric surface is given by:
      σ(m, τ ) = α1 + α2(m − 1)2 + α3(m − 1)3 + α4√τ ,

where m = K/S is the moneyness and τ is the time to maturity of the option in years.
```{r Prepare data for volatility surface fitting, message=FALSE, warning=FALSE}
# Use only OTM calls and OTM puts for surface calibration
options_data <- prepare_options_data_otm(
  calls = calls,
  puts = puts,
  spot_price = last_trading_price
)
```

```{r}
# Fit the parametric surface on OTM quotes
fit_surface <- fit_vol_surface_otm(
  options_data = options_data,
  last_volatility = last_volatility
)

alpha_hat <- fit_surface$alpha_hat
print(alpha_hat)
```

```{r}
# 3D visualization: fitted surface + market IV points (OTM)
iv_surface_3d <- plot_vol_surface_3d(
  options_data = options_data,
  alpha_hat = alpha_hat,
  output_path = here("Outputs", "iv_surface_3d.html")
)

iv_surface_3d
```

Re-price the portfolio using the parametric model. Once you have fitted the volatility surface, consider
˜σ(m, τ ) = σ(m, τ ) − (α1 + α4) + VIXT , where VIXT is the VIX at time T . Use this for the pricing at
time T . For the pricing at T + 5, consider ˜σ(m, τ ) = σ(m, τ ) − (α1 + α4) + VIXT +5 where VIXT +5 is
obtained by simulation.

```{r}
# Re-price the portfolio using the fitted parametric surface sigma_hat(m, tau) (no VIX adjustment yet)
surface_prices_sigma_hat_T <- portfolio_surface_prices_sigma_hat(
  portfolio = portfolio,
  alpha_hat = alpha_hat,
  spot_price = last_trading_price,
  rf_structure = rf_structure,
  trading_days_in_year = trading_days_in_year,
  horizon_days = 0
)

surface_prices_sigma_hat_T$portfolio_value
```


```{r}
# Build adjusted volatility at time T for the portfolio (sigma to plug into BMS)
surface_inputs_T <- portfolio_surface_inputs(
  portfolio = portfolio,
  alpha_hat = alpha_hat,
  spot_price = last_trading_price,
  vix_level = last_volatility,
  trading_days_in_year = trading_days_in_year,
  horizon_days = 0
)

surface_inputs_T
```

```{r}
# Vol surface P&L using bivariate Gaussian scenarios (spot + VIX)
pnl_surface_biv <- calculate_pnl_vol_surface(
  portfolio = portfolio,
  alpha_hat = alpha_hat,
  spot_T = last_trading_price,
  vix_T = last_volatility,
  rf_structure_T = rf_structure,
  rf_structure_T_plus_h = rf_structure_forward,
  trading_days_in_year = trading_days_in_year,
  horizon_days = days_ahead,
  spot_scenarios_T_plus_h = scenarios_asset_biv,
  vix_scenarios_T_plus_h = scenarios_vix
)

pnl_distribution_surface_biv <- pnl_surface_biv$pnl_distribution
VaR95_surface_biv <- pnl_surface_biv$VaR95
ES95_surface_biv  <- pnl_surface_biv$ES95

plot_pnl_vol_surface(
  pnl_distribution = pnl_distribution_surface_biv,
  VaR95 = VaR95_surface_biv,
  ES95 = ES95_surface_biv,
  horizon_days = days_ahead,
  caption = "Vol surface model — VIX from bivariate Gaussian (spot + VIX jointly simulated)",
  output_path = here("Outputs", "pnl_distribution_vol_surface_bivariate.png")
)
```

===========================Annexe=================================

```{r visualize the volatility surface, message=FALSE, warning=FALSE}
# Cleaning filters — kept separate so options_data remains intact
IV_MAX        <- 0.8
MONEYNESS_MIN <- 0.7
MONEYNESS_MAX <- 1.3
CALL_M_MAX    <- 1.15
PUT_M_MIN     <- 0.85

options_data_clean <- options_data %>%
  filter(Implied_Volatility < IV_MAX,
         Moneyness > MONEYNESS_MIN,
         Moneyness < MONEYNESS_MAX)

options_data_calls <- options_data_clean %>% filter(Type == "Call", Moneyness < CALL_M_MAX)
options_data_puts  <- options_data_clean %>% filter(Type == "Put",  Moneyness > PUT_M_MIN)

# Shared fitted surface grid
f_make_surface_grid <- function(data) {
  moneyness_grid <- seq(min(data$Moneyness), max(data$Moneyness), length.out = 50)
  tau_grid       <- seq(min(data$Tau),       max(data$Tau),       length.out = 50)
  surface_grid   <- outer(moneyness_grid, tau_grid, function(m, tau) f_vol_surface(alpha_hat, m, tau))
  list(m = moneyness_grid, tau = tau_grid, z = surface_grid)
}

# Shared plot builder
f_vol_surface_plot <- function(data, title) {
  grid <- f_make_surface_grid(data)
  plot_ly() %>%
    add_surface(
      x = grid$m, y = grid$tau, z = t(grid$z),
      opacity    = 0.6,
      colorscale = list(c(0, 1), c("#2C3E50", "#E74C3C")),
      showscale  = FALSE,
      name       = "Fitted Surface"
    ) %>%
    add_trace(
      data = data,
      x = ~Moneyness, y = ~Tau, z = ~Implied_Volatility,
      type = "scatter3d", mode = "markers",
      marker = list(
        size       = 2,
        color      = ~Implied_Volatility,
        colorscale = list(c(0, 1), c("#AEB6BF", "#E74C3C")),
        showscale  = TRUE,
        colorbar   = list(title = "IV")
      ),
      name = "Market IV"
    ) %>%
    layout(
      title = title,
      scene = list(
        xaxis  = list(title = "Moneyness (K/S)"),
        yaxis  = list(title = "Tau (years)"),
        zaxis  = list(title = "Implied Volatility"),
        camera = list(eye = list(x = 1.8, y = -1.8, z = 0.8))
      )
    )
}

vol_surface_calls <- f_vol_surface_plot(options_data_calls, "Implied Volatility Surface — Calls")
vol_surface_puts  <- f_vol_surface_plot(options_data_puts,  "Implied Volatility Surface — Puts")

vol_surface_calls
vol_surface_puts
```
