---
title: "TP2 - Statistical Methods"
author: "Francis Mazzone"
date: "2026-03-10"
output: pdf_document
header-includes:
  - \usepackage{xcolor}
  - \definecolor{matlabOrange}{RGB}{217,85,26}
  - \usepackage{titling}
  - \pretitle{\begin{center}\color{matlabOrange}\Huge}
  - \posttitle{\end{center}}
editor_options: 
  markdown: 
    wrap: 72
---


```{r setup, include= FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = 'center',
  fig.width = 6,
  fig.height = 4  # note: use fig.align not fig_align
)
library(zoo)
library(forecast)
library(quantmod)
library(stats)
library(dplyr)
library(tidyr)
library(PerformanceAnalytics)
library(cvar)
library(MASS)
```

## Data Import
```{r}
source('TP2_func_stats.R') #BS_call, vol_surface, min_dist_vol_surf
load('Market.rda')
sp500 <- Market$sp500
vix <- Market$vix
int_rates <- Market$rf
calls <- Market$calls
puts <- Market$puts
```

## Pricing of a portfolio of options

#### European call options book setup
```{r} 
call_colnames <- colnames(calls)
eur_k <- c(1600, 1650, 1750, 1800)
eur_tau <- c(20, 20, 40, 40) /250 #Use 250 days/year for Time to maturity
eur_S <- rep(sp500[length(sp500)], 4)
eur_IV <- rep(vix[length(vix)], 4)
eur_P <- rep(NA, 4)

#Linear interpolation for rf
rf_20_days <- approx(x = names(int_rates), y = int_rates, xout = 20/360)$y
rf_40_days <- approx(x = names(int_rates), y = int_rates, xout = 40/360)$y

eur_rf <- c(rep(rf_20_days,2), rep(rf_40_days, 2))

eur_calls <- cbind(eur_k, eur_tau, eur_S, eur_IV, eur_rf, eur_P)
eur_calls
```

#### Finding the price of European calls book
```{r}
eur_calls[, "eur_P"] <- BS_call( 
  S     = eur_calls[, "eur_S"],
  K     = eur_calls[, "eur_k"],
  rf    = eur_calls[, "eur_rf"],
  sigma = eur_calls[, "eur_IV"],
  T_to_mat = eur_calls[, "eur_tau"]
)

P_BS <- eur_calls[, "eur_P"]
eur_calls
```

## Scenario 1: One risk driver and univariate Gaussian model

#### 10,000 simulations for one-week ahead underlying price
```{r}
sp500_r_log <- diff(log(sp500))[-1] #NOTE: Daily returns

#Assume iid 
mu_sp500 <- mean(sp500_r_log)
sigma_sp500 <- sd(sp500_r_log)

n_simulations <- 10000

set.seed(30)
r_sim <- rnorm(n_simulations, mean= mu_sp500 * 5, sd = sigma_sp500 * sqrt(5))
price_sim <- as.numeric(sp500[length(sp500)]) * exp(r_sim)

#price_sim contains the 10k simulated 5-day ahead prices of the underlying asset
```

#### P&L distribution calculation of 5 days ahead
```{r}
call_p_sim <- matrix(nrow= 10000, ncol = 4) #10000 simulations w/ 4 call options

#5 days later so new time to maturity + new interpolated rf rate
tau_new <- c(15, 15, 35, 35) /250

rf_15_days <- approx(x = names(int_rates), y = int_rates, xout = 15/360)$y
rf_35_days <- approx(x = names(int_rates), y = int_rates, xout = 35/360)$y
rf_new <- c(rep(rf_15_days,2), rep(rf_35_days, 2))

for (i in 1:length(price_sim)) {
  call_p_sim[i, ] <- BS_call(
    S        = rep(price_sim[i], 4),
    K        = eur_calls[, "eur_k"],
    rf       = rf_new,
    sigma    = eur_calls[, "eur_IV"],
    T_to_mat = tau_new
  )
}

PnL_univar <- rowSums(call_p_sim) - sum(eur_calls[,"eur_P"]) #P&L of each 10k sims

hist(PnL_univar, freq = FALSE, 
     main = "10000 simulated P&L Univariate Gaussian Distribution", 
     xlab = "P&L", ylab = "Density",
     col = "blue")
lines(density(PnL_univar), lw = 3)
```

#### VaR95 and ES95 for single risk driver and univariate gaussian model

```{r}
VaR95_uni_gauss <- cvar::VaR(PnL_univar, loss = 0.05)

ES95_uni_gauss <- cvar::ES(PnL_univar, p_loss = 0.05)

cat("\nThe VaR95 of the simulated P&Ls is", VaR95_uni_gauss, "\n") 
cat("\nThe ES95 of the simulated P&Ls is", ES95_uni_gauss, "\n") 
```


## Scenario 2: Two Risk Drivers and Bivariate Gaussian Model

#### Setup for bivariate Gaussian
```{r}
vix_r_log <- diff(log(vix))[-1]
mu_vix <- mean(vix_r_log)
sigma_vix <- sd(vix_r_log)

mu_vector <- c(mu_sp500, mu_vix)
Cov_matrix <- cov(cbind(sp500_r_log, vix_r_log))

set.seed(30)
r_bivar_sim <- mvrnorm(n = n_simulations,
                           mu = mu_vector,
                           Sigma = Cov_matrix)

price_sim_bivar <- as.numeric(sp500[length(sp500)]) * exp(r_bivar_sim[,"sp500_r_log"])
sigma_sim_bivar <- as.numeric(vix[length(vix)]) * exp(r_bivar_sim[, "vix_r_log"])

call_p_sim_bivar <- matrix(nrow = 10000, ncol = 4)

#We use tau_new and rf_new of univariate case as those variables have not changed for this scenario

for (i in 1:10000) {
  call_p_sim_bivar[i, ] <- BS_call(
    S        = rep(price_sim_bivar[i], 4),
    K        = eur_calls[, "eur_k"],
    rf       = rf_new,
    sigma    = rep(sigma_sim_bivar[i], 4),
    T_to_mat = tau_new
  )
}

PnL_bivar <- rowSums(call_p_sim_bivar) - sum(eur_calls[, "eur_P"])

hist(PnL_bivar, freq = FALSE, 
     main = "10000 simulated P&L Bivariate Gaussian Distribution", 
     xlab = "P&L", ylab = "Density",
     col = "red")
lines(density(PnL_bivar), lw = 3)
```

## Scenario 3: Two risk drivers and copula-marginal model (Student-t and Gaussian copula)

#### Setup: Fitting the marginals w/ DOF = 10 (sp500) and DOF = 5 (vix)
```{r}
fit_t_sp500 <- fitdistr(sp500_r_log, densfun = "t", df = 10)
mu_t_sp500 <- fit_t_sp500$estimate["m"]
sigma_t_sp500 <- fit_t_sp500$estimate["s"]

fit_t_vix <- fitdistr(vix_r_log, densfun = "t", df = 10)
mu_t_vix <-fit_t_vix$estimate["m"]
sigma_t_vix <- fit_t_vix$estimate["s"]
```

#### Copula
```{r}
library(copula)
corr_sp500_vix <- cor(sp500_r_log, vix_r_log)

gauss_copula <- normalCopula(as.numeric(corr_sp500_vix), dim = 2)

set.seed(30)
cop_sim <- rCopula(n = n_simulations, copula= gauss_copula)

#Applying Inverse Transform + rescaling to original data estimates
sim_sp500_t <- qt(cop_sim[,1], df = 10) * sigma_t_sp500 * sqrt(5) + mu_t_sp500 * 5
sim_vix_t   <- qt(cop_sim[,2], df = 5)  * sigma_t_vix   * sqrt(5) + mu_t_vix   * 5

#5 day ahead price 
price_sim_copula <- as.numeric(sp500[length(sp500)]) * exp(sim_sp500_t)
sigma_sim_copula <- as.numeric(vix[length(vix)]) * exp(sim_vix_t)

call_p_sim_copula <- matrix(nrow = 10000, ncol = 4)

for (i in 1:10000) {
  call_p_sim_copula[i, ] <- BS_call(
    S        = rep(price_sim_copula[i], 4),
    K        = eur_calls[, "eur_k"],
    rf       = rf_new,
    sigma    = rep(sigma_sim_copula[i], 4),
    T_to_mat = tau_new
  )
}

PnL_copula <- rowSums(call_p_sim_copula) - sum(eur_calls[, "eur_P"])

hist(PnL_copula, freq = FALSE, 
     main = "10000 simulated P&L Gaussian Copula w/ Student t-marginals", 
     xlab = "P&L", ylab = "Density",
     col = "green")
lines(density(PnL_copula), lw = 3)
```

# Scenario 4: Volatility Surface

#### Setup 
```{r}
all_options <- rbind(calls, puts)

S_current <- as.numeric(sp500[length(sp500)])

m <- all_options[, "K"] / S_current      # moneyness = K/S
tau <- all_options[, "tau"]               # in years
IV_market <- all_options[, "IV"]        
```

#### Optimize Volatility surface
```{r}
alpha_init <- c(mean(IV_market), 0,0,0)    # initial guess for alphas

fit <- optim(par = alpha_init, 
             fn = min_dist_vol_surf,
             #m = m, tau = tau, IV_market = IV_market,
             method = "Nelder-Mead")

alpha_hat <- fit$par
alpha_hat

```
#### Re-pricing of portfolio with volatility surface
```{r}
eur_calls_m <- eur_calls[ ,"eur_k"] / eur_calls[ ,"eur_S"]

eur_calls_IV_surf <- vol_surface(alpha_hat, 
                                 eur_calls_m, 
                                 eur_calls[ ,"eur_tau"]) - 
                                  (alpha_hat[1] + alpha_hat[4]) + 
                                  eur_calls[, "eur_IV"]
  

eur_calls_p_surf <- BS_call( 
  S     = eur_calls[, "eur_S"],
  K     = eur_calls[, "eur_k"],
  rf    = eur_calls[, "eur_rf"],
  sigma = eur_calls_IV_surf,
  T_to_mat = eur_calls[, "eur_tau"]
)

print("European calls prices at time T w/ Volatility Surface fit:")
eur_calls_p_surf
```

#### T+5 simulation w/ volatility surface
```{r}
eur_calls_p_sim_surf <- matrix(nrow = 10000, ncol =4)

####### SEE IF PRICE_SIM_COPULA IS APPROPRIATE OR NOT
for (i in 1:10000) {
  # Moneyness at T+5 uses simulated underlying price
  m_T5 <- eur_calls[, "eur_k"] / price_sim_copula[i]
  
  # Adjusted volatility surface at T+5 using simulated VIX
  sigma_tilde_T5 <- vol_surface(alpha_hat, m_T5, tau_new) - 
                    (alpha_hat[1] + alpha_hat[4]) + 
                    sigma_sim_copula[i]
  
  eur_calls_p_sim_surf[i, ] <- BS_call(
    S        = rep(price_sim_copula[i], 4),
    K        = eur_calls[, "eur_k"],
    rf       = rf_new,
    sigma    = sigma_tilde_T5,
    T_to_mat = tau_new
  )
}

PnL_volsurf <- rowSums(eur_calls_p_sim_surf) - sum(eur_calls_p_surf)

hist(PnL_volsurf, freq = FALSE, 
     main = "10000 simulated P&L Volatility Surface", 
     xlab = "P&L", ylab = "Density",
     col = "purple")
lines(density(PnL_volsurf), lw = 3)

print("Difference between copula and volatility surface:")
all.equal(PnL_copula, PnL_volsurf)
```


## Comparison between Univariate vs Bivariate Gaussian vs Copula distributions:
```{r}
plot(density(PnL_univar, adjust = 1.5), 
     main = "P&L Distribution of all 4 scenarios",
     xlab = "P&L",
     col = "blue",
     lwd = 2,
     xlim = range(c(-250, 400)),
     ylim = c(0, 0.01)
     )

lines(density(PnL_bivar, adjust = 1.5), 
      col = "red", 
      lwd = 2)

lines(density(PnL_copula, adjust = 1.5),
      col = "green",
      lwd = 2)

lines(density(PnL_volsurf, adjust = 1.5),
      col = "purple",
      lwd = 2)

legend("topright", 
       legend = c("Univariate Gaussian", "Bivariate Gaussian", 
                  "Gaussian Copula (Student marginals)", "Volatility Surface"),
       col = c("blue", "red", "green", "purple"),
       cex = 0.6,
       lwd = 2)
```
